### 2018-03-02 submission/manuscript.tex
discussion and journal csl update

Colorectal cancer (CRC) is a growing world-wide health problem in which
the microbiota has been hypothesized to have a role in disease
progression {[}1,2{]}. Numerous studies using murine models of CRC have
shown the importance of both individual microbes {[}3--7{]} and the
overall community {[}8--10{]} in tumorigenesis. Numerous case-control
studies have characterized the microbiota of individuals with colonic
adenomas and carcinomas in an attempt to identify biomarkers of disease
progression {[}6,11--17{]}. Because current CRC screening
recommendations are poorly adhered to due to socioeconomic status, test
invasiveness, and frequency of tests, development and validation of
microbiome-associated biomarkers for CRC progression could further
attempts to develop non-invasive diagnostics {[}18{]}.

Recently, there has been an intense focus on identifying
microbiota-based biomarker yielding a seemingly endless number of
candidate taxa. Some studies point towards mouth-associated genera such
as Fusobacterium, Peptostreptococcus, Parvimonas, and Porphyromonas that
are enriched in people with carcinomas {[}6,11--17{]}. Other studies
have identified members of \emph{Akkermansia}, \emph{Bacteroides},
\emph{Enterococcus}, \emph{Escherichia}, \emph{Klebsiella},
\emph{Mogibacterium}, \emph{Streptococcus}, and \emph{Providencia} are
also associated with carciomas {[}13--15{]}. Additionally,
\emph{Roseburia} has been found in some studies to be more abundant in
people with tumors but in other studies it has been found to be either
less abundant or no different than what is found in subjects with normal
colons {[}14,17,19,20{]}. There are strong results from tissue culture
and murine models that Fusobacterium nucleatum, pks-positive strains of
Escherichia coli, Streptococcus gallolyticus, and an
entertoxin-producing strain of Bacteroides fragilis are important in the
pathogenesis of CRC {[}5,14,21--24{]}. These results point to a
causative role for the microbiota in CRC pathogenesis as well as their
potential as diagnostic biomarkers.

Most studies have focused on identifying biomarkers in patients with
carcinomas but there is a greater clinical need to identify biomarkers
associated with adenomas. Studies focusing on broad scale community
metrics have found that measures such as the total number of Operational
Taxonomic Units (OTUs) are decreased in those with adenomas versus
controls {[}25{]}. Other studies have identified \emph{Acidovorax},
\emph{Bilophila}, \emph{Cloacibacterium}, \emph{Desulfovibrio},
\emph{Helicobacter}, \emph{Lactobacillus}, \emph{Lactococcus},
\emph{Mogibacterium}, and \emph{Pseudomonas} to be enriched in those
with adenomas {[}25--27{]}. There are few genera that are enriched in
patients with adenoma or carcinoma tumors.

Confirming some of these previous findings, a recent meta-analysis found
that 16S rRNA gene sequences from members of the \emph{Akkermansia},
\emph{Fusobacterium}, and \emph{Parvimonas} were fecal biomarkers for
the presence of carcinomas {[}28{]}. Contrary to previous studies they
found sequences similar to members of \emph{Lactobacillus} and
\emph{Ruminococcus} to be enriched in patients with adenoma or carcinoma
relative to those with normal colons {[}12,15,16{]}. In addition, they
found 16S rRNA gene sequences from members of \emph{Haemophilus},
\emph{Methanosphaera}, \emph{Prevotella}, \emph{Succinovibrio} were
enriched in patients with adenoma and \emph{Pantoea} were enriched in
patients with carcinomas. Although this meta-analysis was helpful for
distilling a large number of possible biomarkers, the aggregate number
of samples included in the analysis (n = 509) was smaller than several
larger case-control studies that have since been published {[}12,27{]}

Here we provide an updated meta-analysis using 16S rRNA gene sequence
data from both feces (n = 1737) and colon tissue (492 samples from 350
individuals) from 14 studies {[}11--17,19,20,23,25--27,29{]} {[}Table 1
\& 2{]}. We expand both the breadth and scope of the previous
meta-analysis to investigate whether biomarkers describing the bacterial
community or specific members of the community can more accurately
classify patients as having adenoma or carcinoma. Our results suggest
that the bacterial community changes as disease severity worsens and
that that a subset of the microbial community can be used to diagnose
the presence of carcinoma.

@@

Select subset community you can do as well as OTUs or genera when using
all of the information and that these models perform well when trained
on one cohort and tested on another.

Why didn't the whole community work better? Communities are patchy and
higher level taxonomic information pools some of that patchiness There
isn't one single bug associated with disease -- why ? possibly because
multiple bugs can carry out the same function (e.g.~inflammation). Tell
some examples (Esch Fuso, pepto) Also, interesting that the loss of
bacteria can be associated with disease. These appear to be commonly
thought as beneficial bacteria that ferment to produce SCFAs

Adenomas suck. Which is kind of at odds with Baxter, but Baxter looked
at lesions and also included FIT Also, variation in how adenomas are
defined -- small polyps vs larger neopplasias (e.g.~baxter), but we
couldn't do this for the current study because we didn't have that level
of detail and most studies are small This isn't such a novel result
(Nich Chia) Stool may not be the best place to look for early stage
adenomas

Tissue vs.~Stool Not much overlap between stool and tissue biomarkers
Stool for diagnostic -- does as well as tissue for classifying
carcinomas Carcinoma result surprising (e.g.~lack of Fuso) -- largely
because small studies?

Missing microbes ETBfrag, Strepto, etc. B -- is a lot of things that we
might lose -- good and bad; mucosal, right sided tumors? We might not
see with stool

@@

Looking forward Potential for using microbiome as a diagnostic tool
We're excited that we were able to validate a set of biomarkes across
multiple studies Found a phenotype we can relate to the microbiome, it's
replicable and it's strong

\emph{Targeting the identification of tumor microbial biomarkers within
stool seems logical since it offers an easy and cost-effective way to
stratify risk of disease. The current gold standard for diagnosis, a
colonoscopy, can be time-consuming and is not without risk of
complications. Although stool represents an easy and less invasive way
to assess risk, it is not clear how well this sample reflects adenoma-
and carcinoma- associated microbial communities. Some studies have tried
to assess this in health and disease but are limited by their sample
size {[}17,30{]}. Sampling the microbiota directly associated with colon
tissue may provide clearer answers but is not without their own
limitations. After the colonoscopy bowel prep the bacterial community
sampled may reflect the better adhered microbiota versus the resident
community. Additionally, these samples contain more host DNA,
potentially limiting the types of analysis that can be done. It is well
known that low biomass samples can be very difficult to work with and
results can be study dependent due to the randomness of contamination
{[}31{]}.}

Our study identifies clear but small differences in diversity at the
community level and larger differences for individual genera, present in
patients with tumors versus controls {[}Figure 1-3{]}. Although there
was a step-wise decrease in diversity as disease progressed from control
to adenoma to carcinoma, this did not translate into large effect size
increases in OR for either adenoma or carcinoma tumors. Even though
mouth-associated genera increased individaul's OR of having a carcinoma
for certain sample types, they did not consistently increase the OR of
having an adenoma. By using these taxa that had significant ORs after
multiple comparison correction we found that we could classify
indviduals with either adenoma or carcinoma as well as models that use
either all genera or all OTUs. Finally, many studies were individually
under powered to be able to reject the null hypothesis and this could
one reason only the comparison between control and carcinoma individuals
for stool samples had relible detectable differences.

The data presented herein support the importance of specific taxa for
carcinoma, but not necessarily adenoma, tumor formation. The results
that we have presented show that the significant OR taxa model and both
the full genera and OTU models, for indviduals with carcinoma, had
similar AUCs {[}Figure 2 \& 3{]}. This suggests that an interplay
between a select number of potentially protective and exacerbating
microbes within the GI community could be crucial for carcinoma
formation. Importantly, it suggests that there may be key members of the
GI community that should be studied further to potentially help reduce
the risk of carcinoma tumor formation. Conversely, using the present
data, it is clear that new approaches may be needed to identify members
of the community associated with adenoma tumors. Regardless of sample
type and whether a full genera- or OTU-based model was used, our Random
Forest models consistently performed poorly. Yet, the step-wise decrease
in diversity suggests that the adenoma-associated community is not
normal but has changed subtly. This change in diversity, at this early
stage of disease, could be focal to the adenoma itself. How the host
interacts with these subtle changes at early stages of the disease could
be what leads to a thoroughly dysfunctional community that is supportive
of tumorgenesis.

For the full genera- and OTU-based models within stool, common GI
microbes were most consistently present in the top 10 genera or OTUs
across studies {[}Figure 4{]}. Changes in \emph{Bacteroides},
\emph{Ruminococcaceae}, \emph{Ruminococcus}, and \emph{Roseburia} were
consistently found to be in the top 10 most important variables across
the different studies for both individuals with adenoma and carcinoma
{[}Figure 4{]}. These data suggest that whether the non-resident
bacterium is \emph{Fusobacteria} or \emph{Peptostreptococcus} may not be
as important as how these bacteria interact with the changing resident
community. Based on these observations, it is possible to hypothesize
that small changes in community structure lead to new niches in which
any one of the mouth-associated or general inflammatory genera can gain
a foothold, exacerbating the initial changes in community and
facilitating the transition from adenoma to carcinoma stage of disease.

The colon tissue-based studies did not provide a clearer understanding
of how the microbiota may be associated with tumors. Generally, the full
OTU-based models of unmatched and matched colon tissue samples were
concordant with stool samples showing that GI resident microbes were the
most prevalent in the top 10 most important variables across study
{[}Figure S4E \& F{]}. Unlike in stool, \emph{Fusobacterium} was the
only mouth-associated bacteria consistently present in the top 10 most
important variables of the full carcinoma stage models {[}Figure S4B-C
\& E-F{]}. The majority of the colon tissue-based results seem to be
study specific with many of the top 10 taxa being present only in a
single study. Additionally, the presence of genera associated with
contamination, within the top 10 most important variables for the genera
and OTU models is worrying. The low bacterial biomass of tissue samples
coupled with potential contamination could explain why these results
seem to be more sporadic than the stool results.

One important caveat to this study is that even though genera associated
with certain species such as \emph{Bacteroides fragilis} and
\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} were not
identified, it does not necessarily mean that these specific species are
not important in human CRC {[}22,24{]}. Since we are limited in our
aggregation of the data to the genus level, it is not possible to
clearly delineate which species are contributing to overall disease
progression. Our observations are not inconsistent with the previous
literature on either \emph{Bacteroides fragilis} or \emph{Streptococcus
gallolyticus} subsp. \emph{gallolyticus}. As an example, the stool-based
full community models consistently identified the genus
\emph{Bacteroides}, as well as OTUs that classified as
\emph{Bacteroides}, to be important model components across studies.
This suggests that even though \emph{Bacteroides} may not increase the
OR of individuals having an adenoma or carcinoma and may not vary in
relative abundance, like \emph{Fusobacterium}, it is still important in
CRC. Additionally, \emph{Streptococcus gallolyticus} subsp.
\emph{gallolyticus} is a mouth-associated microbe, and the results from
this study suggest that regardless of sample type, mouth-associated
genera are commonly associated with an increased OR for individuals to
have a carcinoma tumor.

The associations between the microbiota and individuals with adenoma
tumors are inconclusive, in part, because many studies may not be
powered effectively to observe small effect sizes. None of the studies
analyzed were properly powered to detect a 10\% or lower change between
cases and controls. The results within our meta-analysis suggest that a
small effect size may well be the scope in which differences
consistently occur between controls and those with adenomas. Future
studies investigating adenoma tumors and the microbiota need to take
power into consideration to reproducibly study whether the microbiota
contributes to adenoma formation. In contrast to adenoma stage of
disease, our observations suggest that most studies analyzed have
sufficient power to detect many changes in the carcinoma-associated
microbiota because of large effect size differences between cases and
controls {[}Figure 5{]}.

\subsection{Conclusion}\label{conclusion}

By aggregating together a large collection of studies analyzing both
fecal and colon tissue samples, we are able to provide evidence
supporting the importance of the bacterial community in colorectal
tumors. The data presented here suggests that mouth-associated microbes
can gain a foothold within the colon and are commonly associated with
the greatest OR of individuals having a carcinoma. Conversely, no
conclusive signal with these mouth-associated microbes could be detected
for individuals with an adenoma. Our observations also highlight the
importance of power and sample number considerations when investigating
the microbiota and adenoma tumors due to possible subtle changes in the
community. Overall, the associations between the microbiota and
individuals with carcinomas were much stronger than with those with
adenomas.

@@

\textbf{\emph{Obtaining Data Sets:}} The studies used for this
meta-analysis were identified through the review articles written by
Keku, \emph{et al.} and Vogtmann, \emph{et al.} {[}32,33{]} and
additional studies not mentioned in the reviews were obtained based on
the authors' knowledge of the literature. Studies that used tissue or
feces as their sample source for 454 or Illumina 16S rRNA gene
sequencing analysis and had data sets with sequences available for
analysis were included. Some studies were excluded because they did not
have publicly available sequences or did not have metadata in which the
authors were able to share. After these filtering steps, the following
studies remained: Ahn, \emph{et al.} {[}11{]}, Baxter, \emph{et al.}
{[}12{]}, Brim, \emph{et al.} {[}29{]}, Burns, \emph{et al.} {[}15{]},
Chen, \emph{et al.} {[}13{]}, Dejea, \emph{et al.} {[}20{]}, Flemer,
\emph{et al.} {[}17{]}, Geng, \emph{et al.} {[}19{]}, Hale, \emph{et
al.} {[}27{]}, Kostic, \emph{et al.} {[}34{]}, Lu, \emph{et al.}
{[}26{]}, Sanapareddy, \emph{et al.} {[}25{]}, Wang, \emph{et al.}
{[}14{]}, Weir, \emph{et al.} {[}23{]}, and Zeller, \emph{et al.}
{[}16{]}. The Zackular {[}35{]} study was not included because the 90
individuals analyzed within the study are contained within the larger
Baxter study {[}12{]}. After sequence processing, all the case samples
for the Kostic study had 100 or less sequences remaining and was
excluded, leaving a total of 14 studies that analysis could be completed
on.

@@

\textbf{\emph{Sequence Processing:}} For the majority of studies, raw
sequences were downloaded from the Sequence Read Archive (SRA)
(\url{ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/})
and metadata were obtained by searching the respective accession number
of the study at the following website:
\url{http://www.ncbi.nlm.nih.gov/Traces/study/}. Of the studies that did
not have sequences and metadata on the SRA, data was obtained from DBGap
(n = 1, {[}11{]}) and directly from the authors (n = 4,
{[}17,23,25,27{]}). Each study was processed using the mothur (v1.39.3)
software program {[}36{]} and quality filtering utilized the default
methods for both 454 and Illumina based sequencing. If it was not
possible to use the defaults, the stated quality cut-offs, from the
study itself, were used instead. Sequences that were made up of an
artificial combination of two or more different sequences and commonly
known as chimeras were identified and removed using VSEARCH {[}37{]}
before \emph{de novo} OTU clustering at 97\% similarity was completed
using the OptiClust algorithm {[}38{]}.

\textbf{\emph{Study Analysis Overview:}} OTU richness, evenness, and
Shannon diversity were first assessed for differences between controls,
adenoma tumors, and carcinoma turmors using both linear mixed-effect
models and ORs. For each individual study the Bray-Curtis index was used
to assess differences between control-adenoma and control-carcinoma
individuals. Next, all common genera were assessed for differences in
ORs for individuals having an adenoma or carcinoma and corrected for
multiple comparisons using the Benjamini-Hochberg method {[}39{]}. We
then built Random Forest models based on all genera, all OTUs, or
significant OR taxa (only using taxa still significant after multiple
comparison correction). For both the full genera and significant OR
taxa, models were trained on one study then tested on the remaining
studies using genera-based relative abundances. The OTU-based models
were built using OTU level data and a 10-fold CV over 100 different
iterations, based on random 80/20 splitting of the data, was used to
generate a range of expected AUCs. This process was repeated for every
study in the meta-analysis. Comparisons of the initial trained model
AUCs for the full genera and significant OR taxa were made to the mean
AUC generated from the 100 different 10-fold CV runs of the respective
OTU-based model. For comparisons in which only control versus adenoma
individuals were made, the carcinoma individuals were excluded from each
respective study. Similarly, for comparisons in which control versus
carcinoma individuals were made the adenoma individuals were excluded
from each respective study. For all analysis completed fecal and tissue
samples were kept separate. Within the tissue groups the data were
further divided between samples from the same individual (matched) and
those from different individuals (unmatched).

@@

\textbf{\emph{Statistical Analysis:}} All statistical analysis after
sequence processing utilized the R (v3.4.3) software package {[}40{]}.
For OTU richness, evenness, and Shannon diversity analysis, values were
power transformed using the rcompanion (v1.11.1) package {[}41{]} and
then Z-score normalized using the car (v2.1.6) package {[}42{]}. Testing
for OTU richness, evenness, and Shannon diversity differences utilized
linear mixed-effect models created using the lme4 (v1.1.15) package
{[}43{]} to correct for study, repeat sampling of individuals (tissue
only), and 16S hyper-variable region used. Odds ratios (OR) were
analyzed using both the epiR (v0.9.93) and metafor (v2.0.0) packages
{[}44,45{]} by assessing how many individuals with and without disease
were above and below the overall median value within each specific
study. OR significance testing utilized the chi-squared test. Diversity
differences measured by the Bray-Curtis index utilized the creation of
distance matrix and testing with PERMANOVA executed with the vegan
(v2.4.5) package {[}46{]}. Random Forest models were built using both
the caret (v6.0.78) and randomForest (v4.6.12) packages {[}47,48{]}. All
figures were created using both ggplot2 (v2.2.1) and gridExtra (v2.3)
packages {[}49,50{]}.

@@

\hypertarget{refs}{}
\hypertarget{ref-siegel_cancer_2016}{}
1. Siegel RL, Miller KD, Jemal A. Cancer statistics, 2016. CA: a cancer
journal for clinicians. 2016;66:7--30.

\hypertarget{ref-flynn_metabolic_2016}{}
2. Flynn KJ, Baxter NT, Schloss PD. Metabolic and Community Synergy of
Oral Bacteria in Colorectal Cancer. mSphere. 2016;1.

\hypertarget{ref-goodwin_polyamine_2011}{}
3. Goodwin AC, Destefano Shields CE, Wu S, Huso DL, Wu X, Murray-Stewart
TR, et al. Polyamine catabolism contributes to enterotoxigenic
Bacteroides fragilis-induced colon tumorigenesis. Proceedings of the
National Academy of Sciences of the United States of America.
2011;108:15354--9.

\hypertarget{ref-abed_fap2_2016}{}
4. Abed J, Emgård JEM, Zamir G, Faroja M, Almogy G, Grenov A, et al.
Fap2 Mediates Fusobacterium nucleatum Colorectal Adenocarcinoma
Enrichment by Binding to Tumor-Expressed Gal-GalNAc. Cell Host \&
Microbe. 2016;20:215--25.

\hypertarget{ref-arthur_intestinal_2012}{}
5. Arthur JC, Perez-Chanona E, Mühlbauer M, Tomkovich S, Uronis JM, Fan
T-J, et al. Intestinal inflammation targets cancer-inducing activity of
the microbiota. Science (New York, NY). 2012;338:120--3.

\hypertarget{ref-kostic_fusobacterium_2013}{}
6. Kostic AD, Chun E, Robertson L, Glickman JN, Gallini CA, Michaud M,
et al. Fusobacterium nucleatum potentiates intestinal tumorigenesis and
modulates the tumor-immune microenvironment. Cell Host \& Microbe.
2013;14:207--15.

>>>>

Colorectal cancer (CRC) is a growing world-wide health problem in which
the microbiota has been hypothesized to have a role in disease
progression (1, 2). Numerous studies using murine models of CRC have
shown the importance of both individual microbes (3--7) and the overall
community (8--10) in tumorigenesis. Numerous case-control studies have
characterized the microbiota of individuals with colonic adenomas and
carcinomas in an attempt to identify biomarkers of disease progression
(6, 11--17). Because current CRC screening recommendations are poorly
adhered to due to socioeconomic status, test invasiveness, and frequency
of tests, development and validation of microbiome-associated biomarkers
for CRC progression could further attempts to develop non-invasive
diagnostics (18).

Recently, there has been an intense focus on identifying
microbiota-based biomarker yielding a seemingly endless number of
candidate taxa. Some studies point towards mouth-associated genera such
as Fusobacterium, Peptostreptococcus, Parvimonas, and Porphyromonas that
are enriched in people with carcinomas (6, 11--17). Other studies have
identified members of \emph{Akkermansia}, \emph{Bacteroides},
\emph{Enterococcus}, \emph{Escherichia}, \emph{Klebsiella},
\emph{Mogibacterium}, \emph{Streptococcus}, and \emph{Providencia} are
also associated with carciomas (13--15). Additionally, \emph{Roseburia}
has been found in some studies to be more abundant in people with tumors
but in other studies it has been found to be either less abundant or no
different than what is found in subjects with normal colons (14, 17, 19,
20). There are strong results from tissue culture and murine models that
Fusobacterium nucleatum, pks-positive strains of Escherichia coli,
Streptococcus gallolyticus, and an entertoxin-producing strain of
Bacteroides fragilis are important in the pathogenesis of CRC (5, 14,
21--24). These results point to a causative role for the microbiota in
CRC pathogenesis as well as their potential as diagnostic biomarkers.

Most studies have focused on identifying biomarkers in patients with
carcinomas but there is a greater clinical need to identify biomarkers
associated with adenomas. Studies focusing on broad scale community
metrics have found that measures such as the total number of Operational
Taxonomic Units (OTUs) are decreased in those with adenomas versus
controls (25). Other studies have identified \emph{Acidovorax},
\emph{Bilophila}, \emph{Cloacibacterium}, \emph{Desulfovibrio},
\emph{Helicobacter}, \emph{Lactobacillus}, \emph{Lactococcus},
\emph{Mogibacterium}, and \emph{Pseudomonas} to be enriched in those
with adenomas (25--27). There are few genera that are enriched in
patients with adenoma or carcinoma tumors.

Confirming some of these previous findings, a recent meta-analysis found
that 16S rRNA gene sequences from members of the \emph{Akkermansia},
\emph{Fusobacterium}, and \emph{Parvimonas} were fecal biomarkers for
the presence of carcinomas (28). Contrary to previous studies they found
sequences similar to members of \emph{Lactobacillus} and
\emph{Ruminococcus} to be enriched in patients with adenoma or carcinoma
relative to those with normal colons (12, 15, 16). In addition, they
found 16S rRNA gene sequences from members of \emph{Haemophilus},
\emph{Methanosphaera}, \emph{Prevotella}, \emph{Succinovibrio} were
enriched in patients with adenoma and \emph{Pantoea} were enriched in
patients with carcinomas. Although this meta-analysis was helpful for
distilling a large number of possible biomarkers, the aggregate number
of samples included in the analysis (n = 509) was smaller than several
larger case-control studies that have since been published (12, 27)

Here we provide an updated meta-analysis using 16S rRNA gene sequence
data from both feces (n = 1737) and colon tissue (492 samples from 350
individuals) from 14 studies (11--17, 19, 20, 23, 25--27, 29) {[}Table 1
\& 2{]}. We expand both the breadth and scope of the previous
meta-analysis to investigate whether biomarkers describing the bacterial
community or specific members of the community can more accurately
classify patients as having adenoma or carcinoma. Our results suggest
that the bacterial community changes as disease severity worsens and
that that a subset of the microbial community can be used to diagnose
the presence of carcinoma.

@@

Although we expected that the full OTU models would perform the best at
classifying individuals with and without carcinomas, our observations
suggest that both the full and significant OR taxa models performed
equally well. These results suggest that lower level classification to
species and strain may not add extra useful information with respect to
prediction models. This has been suggested in previous literature where
metagenomics did not perform better than 16S rRNA gene sequencing data
at classifying individuals with normal colons and those with carcinomas
(30). One possible reason as to why lower level classification may not
result in better models is that the communities are patchy and higher
level taxonomic information pools some of this patchiness allowing for
better prediction models. There may also be a fair bit of data
redundancy within models that utilize more of the community. An example
of this redundancy would be when we trained models on one study and
tested it on the other studies and the AUCs of the models created with
the select OR taxa performed as well as full taxa models {[}Figure
6B{]}.

Our observations also suggest that a small collection of taxa can
classify diease as well as full OTU-based models but that these taxa
individually perform quite poorly {[}Figure 3{]}. This result supports
the contention that there might be redundancy of function even amongst
the taxa included in the significant OR models. As an example multiple
different microbes could be similarily stimulating the activation of
inflammatory pathways and by doing so exacerbate disease progression.
Multiple reports within the literature have found that different
bacteria, such as \emph{Escherichia coli} and \emph{Fusobacterium
nucleatum}, can worsen similarily worsen inflammation in mouse models of
tumorigenesis (5, 6, 21). Although the inflammatory taxa were patchy in
their importance and presence across studies those that were not
typically associated with inflammation were consistently important for
every study {[}Figure 6{]}. The loss of these taxa (\emph{Ruminococcus}
and \emph{Clostridium XI} in stool and \emph{Dorea} and \emph{Blautia}
in unmatched tissue) is particularly interesting because many are
commonly thought to be beneficial due to their involvement in production
of short chain fatty acids (31--33).

The adenoma models as a whole performed poorly in classifying
individuals with and without adenomas. This outcome is not inconsistent
with what has been published previously (27, 34). However, the modeling
results are at odds with results obtained in Baxter, \emph{et al.} (12).
There are some major differences between the models generated in this
meta-analysis and what was used in this previous report. First, the
prevoius report's models investigated the classification of lesions
(individuals with adenoma or carcinoma) and not adenoma alone. The
Baxter, \emph{et al.} models also contained Fecal Immunoglobulin Test
data while our meta-analysis models only contained 16S rRNA gene
sequencing data. Although being able to classify individuals with
adenomas is important, the classification of advanced adenomas is a more
clinically meaningful diagnostic tool (i.e.~those that are at high risk
of progressing to a carcinoma). It is possible that we might have been
able to detect differences in the bacterial community if advanced
adenomas were separated from adenomas but that data was not available
for the majority of studies analyzed. It is also possible that the
initial changes to the bacterial community are focal to where the
initial adenoma develops and would not be easily assessed with a fecal
sample.

Although stool represents an easy and less invasive way to assess risk,
it is not clear how well this sample reflects adenoma- and carcinoma-
associated microbial communities. Some studies have tried to assess this
in health and disease but are limited by their sample size (17, 35).
Sampling the microbiota directly associated with colon tissue may
provide clearer answers but the colon tissue-based studies did not
provide a clearer understanding of how the microbiota may be associated
with tumors. Generally, the full OTU-based models of unmatched and
matched colon tissue samples were concordant with stool samples showing
that GI resident microbes were the most prevalent in the top 10 most
important variables across study {[}Figure S3{]}. \emph{Fusobacterium}
was not identified consistently across studies and this could be due to
both a small number of studies and a small sample size within these
studies. Additionally, the majority of the colon tissue-based results
were study specific with many of the top 10 taxa being present only in a
single study. The presence of genera associated with contamination (36),
within the top 10 most important variables for the genera and OTU models
is worrying (e.g. \emph{Novosphingobium}, \emph{Acidobacteria Gp2},
\emph{Sphingomonas}, etc.). The low bacterial biomass of tissue samples
coupled with potential contamination and small sample sizes could
explain why these results seem to be more sporadic than the stool
results.

One important caveat to this study is that even though genera associated
with certain species such as \emph{Bacteroides fragilis} and
\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} were not
identified, it does not necessarily mean that these specific species are
not important in human CRC (22, 24). There are reports that
\emph{Bacteroides fragilis}, positive for the enterotoxigenic gene, are
found at specific locations along the colon but the samples we were able
to use in this meta-analysis could not identify these types of
differences (37). Additionally, since we are limited in our aggregation
of the data to the genus level, it is not possible to clearly delineate
which species are contributing to overall disease progression. Our
observations are not inconsistent with the previous literature on either
\emph{Bacteroides fragilis} or \emph{Streptococcus gallolyticus} subsp.
\emph{gallolyticus}. As an example, the stool-based full community
models consistently identified the genus \emph{Bacteroides}, as well as
OTUs that classified as \emph{Bacteroides}, to be important model
components across studies. This suggests that even though
\emph{Bacteroides} may not increase the OR of individuals having an
adenoma or carcinoma and may not vary in relative abundance, like
\emph{Fusobacterium}, it is still important in CRC. Additionally,
\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} is a
mouth-associated microbe, and the results from this study suggest that
regardless of sample type, mouth-associated genera are commonly
associated with an increased OR for individuals to have a carcinoma
tumor.

@@

By aggregating together a large collection of studies analyzing both
fecal and colon tissue samples, we are able to provide evidence
supporting the importance of the bacterial community in carcinoma
tumors. Although further validation of the biomarkers presented here
need to be undertaken, the replciability of the AUC of a specific
collection of taxa across multiple studies suggests a strong potential
for the use of the microbiota as a risk stratification tool for
individuals with carcinomas.

@@

\textbf{\emph{Obtaining Data Sets:}} The studies used for this
meta-analysis were identified through the review articles written by
Keku, \emph{et al.} and Vogtmann, \emph{et al.} (38, 39) and additional
studies not mentioned in the reviews were obtained based on the authors'
knowledge of the literature. Studies that used tissue or feces as their
sample source for 454 or Illumina 16S rRNA gene sequencing analysis and
had data sets with sequences available for analysis were included. Some
studies were excluded because they did not have publicly available
sequences or did not have metadata in which the authors were able to
share. After these filtering steps, the following studies remained: Ahn,
\emph{et al.} (11), Baxter, \emph{et al.} (12), Brim, \emph{et al.}
(29), Burns, \emph{et al.} (15), Chen, \emph{et al.} (13), Dejea,
\emph{et al.} (20), Flemer, \emph{et al.} (17), Geng, \emph{et al.}
(19), Hale, \emph{et al.} (27), Kostic, \emph{et al.} (40), Lu, \emph{et
al.} (26), Sanapareddy, \emph{et al.} (25), Wang, \emph{et al.} (14),
Weir, \emph{et al.} (23), and Zeller, \emph{et al.} (16). The Zackular
(41) study was not included because the 90 individuals analyzed within
the study are contained within the larger Baxter study (12). After
sequence processing, all the case samples for the Kostic study had 100
or less sequences remaining and was excluded, leaving a total of 14
studies that analysis could be completed on.

@@

\textbf{\emph{Sequence Processing:}} For the majority of studies, raw
sequences were downloaded from the Sequence Read Archive (SRA)
(\url{ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/})
and metadata were obtained by searching the respective accession number
of the study at the following website:
\url{http://www.ncbi.nlm.nih.gov/Traces/study/}. Of the studies that did
not have sequences and metadata on the SRA, data was obtained from DBGap
(n = 1, (11)) and directly from the authors (n = 4, (17, 23, 25, 27)).
Each study was processed using the mothur (v1.39.3) software program
(42) and quality filtering utilized the default methods for both 454 and
Illumina based sequencing. If it was not possible to use the defaults,
the stated quality cut-offs, from the study itself, were used instead.
Sequences that were made up of an artificial combination of two or more
different sequences and commonly known as chimeras were identified and
removed using VSEARCH (43) before \emph{de novo} OTU clustering at 97\%
similarity was completed using the OptiClust algorithm (44).

\textbf{\emph{Study Analysis Overview:}} OTU richness, evenness, and
Shannon diversity were first assessed for differences between controls,
adenoma tumors, and carcinoma turmors using both linear mixed-effect
models and ORs. For each individual study the Bray-Curtis index was used
to assess differences between control-adenoma and control-carcinoma
individuals. Next, all common genera were assessed for differences in
ORs for individuals having an adenoma or carcinoma and corrected for
multiple comparisons using the Benjamini-Hochberg method (45). We then
built Random Forest models based on all genera, all OTUs, or significant
OR taxa (only using taxa still significant after multiple comparison
correction). For both the full genera and significant OR taxa, models
were trained on one study then tested on the remaining studies using
genera-based relative abundances. The OTU-based models were built using
OTU level data and a 10-fold CV over 100 different iterations, based on
random 80/20 splitting of the data, was used to generate a range of
expected AUCs. This process was repeated for every study in the
meta-analysis. Comparisons of the initial trained model AUCs for the
full genera and significant OR taxa were made to the mean AUC generated
from the 100 different 10-fold CV runs of the respective OTU-based
model. For comparisons in which only control versus adenoma individuals
were made, the carcinoma individuals were excluded from each respective
study. Similarly, for comparisons in which control versus carcinoma
individuals were made the adenoma individuals were excluded from each
respective study. For all analysis completed fecal and tissue samples
were kept separate. Within the tissue groups the data were further
divided between samples from the same individual (matched) and those
from different individuals (unmatched).

@@

\textbf{\emph{Statistical Analysis:}} All statistical analysis after
sequence processing utilized the R (v3.4.3) software package (46). For
OTU richness, evenness, and Shannon diversity analysis, values were
power transformed using the rcompanion (v1.11.1) package (47) and then
Z-score normalized using the car (v2.1.6) package (48). Testing for OTU
richness, evenness, and Shannon diversity differences utilized linear
mixed-effect models created using the lme4 (v1.1.15) package (49) to
correct for study, repeat sampling of individuals (tissue only), and 16S
hyper-variable region used. Odds ratios (OR) were analyzed using both
the epiR (v0.9.93) and metafor (v2.0.0) packages (50, 51) by assessing
how many individuals with and without disease were above and below the
overall median value within each specific study. OR significance testing
utilized the chi-squared test. Diversity differences measured by the
Bray-Curtis index utilized the creation of distance matrix and testing
with PERMANOVA executed with the vegan (v2.4.5) package (52). Random
Forest models were built using both the caret (v6.0.78) and randomForest
(v4.6.12) packages (53, 54). All figures were created using both ggplot2
(v2.2.1) and gridExtra (v2.3) packages (55, 56).

@@

\hypertarget{refs}{}
\hypertarget{ref-siegel_cancer_2016}{}
1. \textbf{Siegel, R. L.}, \textbf{K. D. Miller}, and \textbf{A. Jemal}.
2016. Cancer statistics, 2016. CA: a cancer journal for clinicians
\textbf{66}:7--30.

\hypertarget{ref-flynn_metabolic_2016}{}
2. \textbf{Flynn, K. J.}, \textbf{N. T. Baxter}, and \textbf{P. D.
Schloss}. 2016. Metabolic and Community Synergy of Oral Bacteria in
Colorectal Cancer. mSphere \textbf{1}.

\hypertarget{ref-goodwin_polyamine_2011}{}
3. \textbf{Goodwin, A. C.}, \textbf{C. E. Destefano Shields}, \textbf{S.
Wu}, \textbf{D. L. Huso}, \textbf{X. Wu}, \textbf{T. R. Murray-Stewart},
\textbf{A. Hacker-Prietz}, \textbf{S. Rabizadeh}, \textbf{P. M. Woster},
\textbf{C. L. Sears}, and \textbf{R. A. Casero}. 2011. Polyamine
catabolism contributes to enterotoxigenic Bacteroides fragilis-induced
colon tumorigenesis. Proceedings of the National Academy of Sciences of
the United States of America \textbf{108}:15354--15359.
